draft

### 2025-05-26 draft.tex
sent and anonymized version

\twocolumn[
\icmltitle{cp\_measure: API-first feature extraction for image-based profiling workflows}

@@

\begin{abstract}
Biological image analysis has traditionally focused on measuring specific visual properties of interest for cells or other entities. 
A complementary paradigm gaining increasing traction is image-based profiling - quantifying many distinct visual features to form comprehensive profiles which may reveal hidden patterns in cellular states, drug responses, and disease mechanisms.
While current tools like CellProfiler can generate these feature sets, they pose significant barriers to automated and reproducible analyses, hindering ML workflows. Here we introduce cp\_measure, a Python library that extracts CellProfiler's core measurement capabilities into a modular, API-first tool designed for programmatic feature extraction. We demonstrate that cp\_measure features retain high fidelity with CellProfiler features while enabling seamless integration with the scientific Python ecosystem. Through applications to 3D astrocyte imaging and spatial transcriptomics, we showcase how cp\_measure enables reproducible, automated image-based profiling pipelines that scale effectively for machine learning applications in computational biology.
\end{abstract}

\section{Introduction}
\label{sec:org4c9ba67}
High throughput screening of biological phenomena via complex modalities, such as RNA sequencing, is prohibitively expensive, therefore microscopy is an efficient first step. Through microscopy, modern biologists use a wide array of fluorescence dyes or proteins to observe the location and distribution of cells, organelles, and other components. Nowadays, quantification of biological images is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity distributions.

Image-based profiling--also termed morphological profiling--is a technique of measuring an extensive suite of morphological features for a population of biological objects, such as cells. 
These features are fed into statistical or machine learning methods to identify biologically meaningful patterns.

One of its biggest applications is drug discovery, where scientists leverage microscopy's low acquisition cost and high throughput to accomplish many goals, such as grouping genes by function, identifying chemical compounds that target a protein, and predicting toxicity of drug candidates \citep{sealDecadeSystematicReview2024}.

\begin{figure}[htbp]
\centering
\includesvg[width=.99\linewidth]{./figs/cpmeasure_overview.svg}
\caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object"). It can featurize the pairwise combination of all the available channels (colours) and objects. The resultant matrices represent the entire experiment and can be studied using statistical, machine, and deep learning methods.}
\end{figure}

\subsection{The current state of bioimage analysis}
\label{sec:org8f5b33d}
The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}. It can be used by experimental biologists with scant programming experience, for whom it has been a boon. By contrast, it may not suit the needs of computer-savvy scientists building high throughput pipelines comprised of multiple tools. CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance. For example, complementing CellProfiler with other image analysis tools is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces. Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats.

Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025}. 
%However, these come with constraints: Given that their implementation is completely independent from CellProfiler, they do not produce the same feature set, precluding matching to existing datasets, such as those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}. Moreover, a common problem is that these tools enforce a structure for the input data, such as determined hierarchical file structures, which require significant wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage, neglecting a significant number of cases where local compute suffices.

 However, these approaches present trade-offs for different use cases. 
 Their independent implementation means they generate different feature sets than CellProfiler, which can be advantageous for novel
 applications but complicates direct comparison with established datasets like those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024}. 
Additionally, while all tools necessarily impose input requirements, some enforce rigid organizational constraints—demanding specific directory structures and naming conventions—that require extensive data reorganization before analysis, beyond the standard data type and format specifications.
Some alternatives also target specific deployment scenarios, with ScaleFex designed primarily for cloud environments, limiting utility for researchers preferring local computation.

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
%An alternative to extracting a defined set of engineered features is to instead use a pre-trained deep-learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Compared to engineered features, deep learning approaches may pose a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset, in addition to having access to a GPU. Foundational models aim to lower this barrier, but can fail to generalize well for out-of-distribution data (data which differs sufficiently from the training datasets) \cite{azadFoundationalModelsMedical2023}. Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.  Unlike learned features, whose interpretation requires complex methods like counterfactual techniques or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts, such as the colocalisation of two proteins \cite{garcia-fossaInterpretingImagebasedProfiles2023}.

%In summary, while deep learning-based features have some advantages in some contexts, and while tools to extract engineered features exist, there remains a need for a modular tool that can extract interpretable features while readily integrating with computational workflows to achieve reproducible data analyses, whether locally or in the cloud. We thus believe that there is the need for a readily-integrated feature extraction tool aimed at generalist and high throughput problems not being solved by the current software ecosystem. With all this in mind, we developed cp\_measure.

@@

Several practical considerations favor engineered features in biological imaging. 
First, interpretability remains crucial for biological discovery—researchers need to understand which morphological characteristics drive observed phenotypes. 
While deep learning features can be powerful predictors, their interpretation requires sophisticated techniques that may not provide the mechanistic insights biologists seek \citep{liChallengesOpportunitiesBioimage2023}. 
Engineered features have clear mathematical definitions that directly translate to biological concepts, such as protein colocalization or nuclear morphology changes \cite{garcia-fossaInterpretingImagebasedProfiles2023}.

Second, many biological imaging laboratories lack the GPU resources, large annotated datasets, or machine learning expertise required for custom model development. While foundational models are emerging, they often struggle with the diversity of imaging modalities and experimental conditions \cite{azadFoundationalModelsMedical2023}.

@@

Rather than viewing these approaches as competing alternatives, the field increasingly recognizes their complementary roles. 
This motivates the need for robust, accessible tools that efficiently extract interpretable features while integrating seamlessly with modern computational workflows.
With all this in mind, we developed cp\_measure.

\section{Extracting interpretable features}
\label{sec:org61842b5}
The library cp\_measure branches off the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task. Measurements are defined as a collection of related features, following CellProfiler's nomenclature, and are categorized based on three kinds of input: One object (producing features such as area and eccentricity), one object + one imaging channel (these can be used to extract such measurements as intensity, texture), one object + multiple channels (e.g., for Manders correlations between intensity values across channels).

@@

%First, we demonstrated the validity and capabilities of cp\_measure in cases that would normally require image analysis knowledge and a considerable amount of time. First, we validate cp\_measure features versus CellProfiler results with a subset of the JUMP dataset \citep{chandrasekaranJUMPCellPainting2023}. Then we showcase cases in which cp\_measure is a more practical choice to process microscopy data: First 3D images of astrocytes and then spatial transcriptomics. These use-cases demonstrate its widespread suitability for different types of problems. 
\subsection{Reproducing CellProfiler measurements}
\label{sec:org09b0cd2}
We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images. These correspond to 150 perturbations from the JUMP Cell Painting dataset \citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for each feature. We segmented images to delineate the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction. Next, we applied cp\_measure on these masks and the original images. 
To validate cp\_measure, we compared its output directly with CellProfiler's on identical images and masks. 
For each feature, we computed the $R^2$ value between the two tools' measurements. 
Over 90\% of features showed $R^2$ > 0.9 (Figure \ref{fig:cp_vs_cpmeasure}), indicating near-perfect agreement. 
The few outliers with lower correlations reveal edge cases handled differently by each tool; these will be addressed through comprehensive unit testing.
%Finally, we mapped the features from cp\_measure to CellProfiler and calculated a linear fit for each matched feature, plus the \(R^2\) value, indicating how well the correlation fits a linear slope.
%We find that most features had an $R^2$ correlation larger than 90\% (Figure \ref{fig:cp_vs_cpmeasure}). We examined several individual features, typically finding straight lines demonstrating the recapitulation of features from our implementation. For a small set of features, some data points fall off the diagonal, which hint that some edge cases are treated differently by either tool; we plan to investigate these further as part of a larger project to implement unit testing in CellProfiler. 
These results indicate that cp\_measure is not likely to produce massively different profiles than those calculated by CellProfiler.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
\caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogues. \textbf{Panels A-C.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images. \textbf{Panel D.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
\end{figure}

@@

%To demonstrate its ease of use and the value of interpretable features, we next tested cp\_measure in a classification workflow. We processed 433 three-dimensional (3D) images of astrocyte nuclei containing 831 nuclei \citep{kalininValproicAcidinducedChanges2021}. We preprocessed the profiles following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}. Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired, which corresponds to the extent of maturation of the cells. We achieved 87\% classification accuracy in the testing dataset. Using the image-based profiles, we identified which features distinguish cells based on the time at which the image was taken. Finally, we calculated the SHAP values \citep{sundararajanManyShapleyValues2020} to get a better understanding of the effects of the drugs on the cells.
% Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the SHAP values of a classifier trained on our features. The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover.

We demonstrate cp\_measure's utility through a biologically relevant task: tracking astrocyte maturation over time. 
From 433 3D images containing 831 astrocyte nuclei \citep{kalininValproicAcidinducedChanges2021}, cp\_measure extracted morphological features  that we preprocessed following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
We then trained a Gradient Boosting classifier to predict the day of cell differentiation with 87\% accuracy. 
SHAP analysis \citep{sundararajanManyShapleyValues2020} (\ref{fig:astrocytes}) revealed that nuclear minor axis length was the most predictive feature: nuclei became progressively wider during maturation, a phenotype consistent with known astrocyte biology.

@@

\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
To demonstrate the broad applicability of cp\_measure, we developed a morphology featurizer for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labelled.

% Using CellProfiler on spatial datasets is not convenient due to its enforcement of a nested structure on input data and, since it requires loading entire images into RAM, struggles. Since images from spatial datasets contain thousands of times more cells than conventional microscopy data, usually reaching 500.000 cells per image, the stand-alone nature of cp\_measure allowed us to tailor a workflow to match this data throughput. We used Squidpy to leverage the SpatialData data format \citep{Marconato2024-SpatialData}, enabling us to parallelize the feature extraction with cp\_measure to streaming batches of individual cells across multiple processing cores. This approach enables efficient computation on large-scale datasets, which would be intractable using CellProfiler. Furthermore, this brings morphological profiling into the single-cell -- scverse -- ecosystem \citep{Virshup2023-scverse}, exposing it to the wider single-cell community.

Spatial transcriptomics poses unique computational challenges: images routinely contain 500,000+ cells, far exceeding CellProfiler's memory constraints and rigid data structure requirements. 
cp\_measure's modular design enabled us to build a custom workflow integrated with Squidpy \citep{pallaSquidpyScalableFramework2022} that leverages the SpatialData format \citep{Marconato2024-SpatialData} to process cells in streaming parallel batches. 
This architecture handles massive spatial datasets efficiently while bringing morphological profiling directly into the scverse ecosystem \citep{Virshup2023-scverse}, making these features immediately accessible to the single-cell community.

To illustrate the value of this workflow, we used it to identify morphologically similar regions within two human breast cancer sections profiled with the spatial transcriptomics technology Xenium \citep{10x-Genomics2023-el}. We tessellated hexagonal masks across both images, of which 18469 were found to overlay the tissue. These masks were then featurized with a cp\_measure-based workflow: We first preprocessed the features following standard workflows
\citep{serranoReproducibleImagebasedProfiling2025}, then integrated our two samples using Harmony \citep{Korsunsky2019-Harmony}, and finally obtained the Leiden clusters via Scanpy \citep{wolfSCANPYLargescaleSinglecell2018}.

For each cluster, we extracted the cells within its hexagons and calculated the cell type proportions predicted by gene-expression. A visual comparison of panels A and B in Figure \ref{fig:spatial_omics} shows that morphology-only clustering largely recapitulates the underlying tissue architecture. Panel C reveals that some clusters have distinct cell-type compositions: for example, cluster 2 is dominated by fibroblasts, whereas adipocytes appear almost exclusively in clusters 5-7.

\begin{figure}[htbp]
\centering
\includegraphics[width=.99\linewidth]{./figs/fig_4_spatial.png}
\caption{\label{fig:spatial_omics}{}Morphology reveals the composition of tumour subregions. \textbf{Panel A.} Spatial map of cell types inferred from transcriptomics. \textbf{Panel B.} Morphological subregions obtained by applying Leiden clustering to image-derived feature embeddings. \textbf{Panel C.} Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type (colour legend as shown). The top left quarter of Panel A shows that the morphological approach worked to predict regions while the expression annotations did not.}
\end{figure}

These results demonstrate how cp\_measure can can be used outside traditional microscopy. When applied on extremely large images it produces biologically meaningful morphological clusters that validate and complement transcriptomic phenotypes. By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly add morphology to their spatial omics workflows, unlocking new insights into tissue organization and cell–cell interactions.

\section{Discussion}
\label{sec:orgf37b369}
While point-and-click interfaces open the world of image analysis to many researchers, they are not as effective for computational workflows with no human-in-the-loop. In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks). It enables simpler automated profiling of microscopy data in short scripts and complex pipelines. The modularity it provides facilitates pipelines with better scaling capabilities for high-content microscopy, with or without cloud infrastructure.

The biologically interpretable features provided by cp\_measure complement deep learning features and offer a better mechanistic understanding of the underlying biology. A potential workflow would be to use deep learning features to cluster and then use cp\_measure features from these clusters. As a whole, when used in tandem with generalist image-processing tools, such as Cellpose for segmentation \cite{stringerCellposeGeneralistAlgorithm2021}, machine and deep learning workflows can be streamlined.

While developed for biology, engineered morphological features have proven valuable far beyond their original domain--from analysing microplastics in environmental samples \citep{ideharaExploringNileRed2025} to quality control in manufacturing CITE SOMETHING and agricultural automation \citep{xuDesignNonDestructiveSeed2024}. 
cp\_measure's lightweight, API-first design makes these powerful features accessible to any field working with image data, removing the barriers of biological-specific interfaces.

\section{Future work}
\label{sec:org5cdbb12}
We propose to make cp\_measure an imported dependency for CellProfiler, offering several benefits. It would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements. It would also mean that any changes made in cp\_measure propagate to CellProfiler, benefiting the community for which it was originally developed.

We also plan to develop a comprehensive test suite to guarantee mathematical correctness, which currently CellProfiler itself is lacking. A comprehensive test suite would enable more confident and expedient optimization for the most compute-consuming features, such as granularity, by providing rapid iteration capabilities. Once tests are in place, we could add support for just-in-time compiling and GPUs. We envision that cp\_measure could also be the place to develop and distribute new measurements from the community.

@@

\label{sec:orgdd18dd8}
\subsection{Methods}
\label{sec:orgb3e9382}
\subsubsection{Data and software}
\label{sec:orgbda0ae2}
The codebase for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}. All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://anonymous.4open.science/r/2025_cpmeasure-3FBF}. 
%The datasets we produced for this work are available on Zenodo. The selection of samples from JUMP and the images can be found on \url{https://zenodo.org/records/15390631/latest}. The image-based profiles produced with Cellprofiler and used for the comparisons in Figure 2 can be on \url{https://zenodo.org/records/15505477/latest}.
%

@@

>>>>

\twocolumn[
\icmltitle{LIB\_NAME: API-first feature extraction for image-based profiling workflows}

@@

\begin{abstract}
Biological image analysis has traditionally focused on measuring specific visual properties of interest for cells or other entities. 
A complementary paradigm gaining increasing traction is image-based profiling - quantifying many distinct visual features to form comprehensive profiles which may reveal hidden patterns in cellular states, drug responses, and disease mechanisms.
While current tools like CellProfiler can generate these feature sets, they pose significant barriers to automated and reproducible analyses, hindering machine learning workflows. Here we introduce LIB\_NAME, a Python library that extracts CellProfiler's core measurement capabilities into a modular, API-first tool designed for programmatic feature extraction. We demonstrate that LIB\_NAME features retain high fidelity with CellProfiler features while enabling seamless integration with the scientific Python ecosystem. Through applications to 3D astrocyte imaging and spatial transcriptomics, we showcase how LIB\_NAME enables reproducible, automated image-based profiling pipelines that scale effectively for machine learning applications in computational biology.
\end{abstract}

\section{Introduction}
\label{sec:org4c9ba67}
High throughput screening of biological phenomena via complex modalities, such as RNA sequencing, is prohibitively expensive, therefore microscopy is an efficient first step. Through microscopy, modern biologists use a wide array of fluorescence dyes or proteins to observe the location and distribution of cells, organelles, and other components. Nowadays, quantification of biological images is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity.

Image-based profiling--also termed morphological profiling--is a technique of measuring an extensive suite of morphological features for a population of biological objects, such as cells. 
These features are fed into statistical or machine learning methods to identify biologically meaningful patterns. One of its biggest applications of morphological profiling is drug discovery, where scientists leverage microscopy's low acquisition cost and high throughput to accomplish many goals, such as grouping genes by function, identifying chemical compounds that target a protein, and predicting toxicity of drug candidates \citep{chandrasekaranImagebasedProfilingDrug2021}.

\begin{figure}[htbp]
\centering
\includesvg[width=.99\linewidth]{./figs/cpmeasure_overview.svg}
\caption{\label{fig:overview}LIB\_NAME generates features from images by using information in every region of interest ("object"). It can featurize the pairwise combination of all the available channels (colours) and objects. The resultant matrices represent the entire experiment and can be studied using statistical, machine, and deep learning methods.}
\end{figure}

\subsection{The current state of bioimage analysis}
\label{sec:org8f5b33d}
The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}, providing experimental biologists with limited programming expertise an accessible yet powerful toolset. By contrast, it may not suit the needs of computational biologists building high throughput pipelines comprised of multiple tools. CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles or spatial transcriptomics data, however, certain aspects become a hindrance. For example, complementing CellProfiler with other image analysis tools is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces. Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats.

Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025}, however, these approaches present trade-offs for different use cases. Their independent implementation means they generate different feature sets than CellProfiler, which can be advantageous for novel
 applications but complicates direct comparison with established datasets like those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024}. 
Additionally, while all tools necessarily impose input requirements, some enforce rigid organizational constraints—demanding specific directory structures and naming conventions—that require extensive data reorganization before analysis, beyond the standard data type and format specifications.
Some alternatives also target specific deployment scenarios, with ScaleFex designed primarily for cloud environments, limiting utility for researchers preferring local computation.

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}

@@

Several practical considerations favour engineered features in biological imaging. 
First, interpretability remains crucial for biological discovery—researchers need to understand which morphological characteristics drive observed phenotypes. 
While deep learning features can be powerful predictors, their interpretation requires sophisticated techniques that may not provide the mechanistic insights biologists seek \citep{liChallengesOpportunitiesBioimage2023}. 
Engineered features have clear mathematical definitions that directly translate to biological concepts, such as protein co-localisation or nuclear morphology changes \cite{garcia-fossaInterpretingImagebasedProfiles2023}.

Second, many biological imaging laboratories lack the GPU resources, large annotated datasets, or machine learning expertise required for custom model development. While foundational models are emerging, visual foundational models can be challenged by the diversity of imaging modalities and experimental conditions \cite{azadFoundationalModelsMedical2023}.

@@

Rather than viewing these approaches as competing alternatives, the field increasingly recognizes their complementary roles. 
This motivates the need for robust, accessible tools that efficiently extract interpretable features while integrating seamlessly with modern computational workflows.
With all this in mind, we developed LIB\_NAME.

\section{Extracting interpretable features}
\label{sec:org61842b5}
The library LIB\_NAME branches off the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task. Measurements are defined as a collection of related features, following CellProfiler's nomenclature, and are categorized based on three kinds of input: One object (producing features such as area and eccentricity), one object + one imaging channel (these can be used to extract such measurements as intensity, texture), one object + multiple channels (e.g., for Manders correlations between intensity values across channels). A visual representation of this can be found in Figure~\ref{fig:overview}.

@@

\subsection{Reproducing CellProfiler measurements}
\label{sec:org09b0cd2}
We first tested whether LIB\_NAME features match the original CellProfiler features, for a representative set of 300 images. These correspond to 150 perturbations from the JUMP Cell Painting dataset \citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for each feature. We segmented images to delineate the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to LIB\_NAME for feature extraction. Next, we applied LIB\_NAME on these masks and the original images. 
To validate LIB\_NAME, we compared its output directly with CellProfiler's on identical images and masks. 
For each feature, we computed the $R^2$ value between the two tools' measurements. 
Over 95\% of features showed $R^2$ > 0.9 (Figure \ref{fig:cp_vs_cpmeasure}), indicating near-perfect agreement. 
The few outliers with lower correlations reveal edge cases handled differently by each tool; these will be addressed through comprehensive unit testing. These results indicate that LIB\_NAME is not likely to produce massively different profiles than those calculated by CellProfiler.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
\caption{\label{fig:cp_vs_cpmeasure}LIB\_NAME features match their CellProfiler analogues. \textbf{Panels A-C.} Representative examples comparing CellProfiler feature values (x-axis) to LIB\_NAME's (y-axis), generated using matching pairs of masks and images. \textbf{Panel D.} \(R^2\) value of a linear fit for each individual feature, comparing LIB\_NAME to CellProfiler.}
\end{figure}

@@

We demonstrate LIB\_NAME's utility through a biologically relevant task: tracking astrocyte maturation over time. 
From 433 3D images containing 831 astrocyte nuclei \citep{kalininValproicAcidinducedChanges2021}, LIB\_NAME extracted morphological features that we preprocessed following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
We then trained a gradient-boosting classifier to predict the day of cell differentiation which it was able to do with 86\% accuracy.

SHAP analysis \citep{sundararajanManyShapleyValues2020} (Figure\ref{fig:astrocytes}) revealed that nuclear minor axis length was the most predictive feature: nuclei became progressively wider during maturation, a phenotype consistent with known astrocyte biology.

@@

\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
To demonstrate the broad applicability of LIB\_NAME, we developed a morphology featurizer for the popular spatial omics analysis toolbox Squidpy~\citep{pallaSquidpyScalableFramework2022}.
Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labelled.

Spatial transcriptomics poses unique computational challenges: images routinely contain 500,000+ cells, far exceeding CellProfiler's memory constraints and rigid data structure requirements. 
LIB\_NAME's modular design enabled us to build a custom workflow integrated with Squidpy~\citep{pallaSquidpyScalableFramework2022} that leverages the SpatialData format \citep{Marconato2024-SpatialData} to process cells in streaming parallel batches. 
This architecture handles massive spatial datasets efficiently while bringing morphological profiling directly into the scverse ecosystem \citep{Virshup2023-scverse}, making these features immediately accessible to the single-cell community.

\\

To illustrate the value of this workflow, we used it to identify morphologically similar regions within two human breast cancer sections profiled with the spatial transcriptomics technology Xenium \citep{10x-Genomics2023-el}. We generated a pattern of tessellating hexagonal masks across both images, of which 18469 were found to overlay the tissue. These masks were then featurized with a LIB\_NAME-based workflow: We first preprocessed the features following standard workflows
\citep{serranoReproducibleImagebasedProfiling2025}, then removed batch effects between the samples using Harmony \citep{Korsunsky2019-Harmony}, and finally obtained the Leiden clusters via Scanpy \citep{wolfSCANPYLargescaleSinglecell2018, Traag2019-fr}.

For each cluster, we extracted the cells within its hexagons and calculated the respective cell type proportions. A visual comparison of panels A and B in Figure~\ref{fig:spatial_omics} shows that morphology-only clustering largely recapitulates the underlying tissue architecture. Panel C reveals that some clusters have distinct cell-type compositions: For example, cluster~2 is dominated by fibroblasts, whereas adipocytes appear almost exclusively in clusters 5-7. Furthermore, comparing clusters~0 and 1 suggests intra-tumour differences, which could be correlated to immune-hotness (see Figure \ref{fig:spatial_immunehot}).

\begin{figure}[htbp]
\centering
\includegraphics[width=.99\linewidth]{./figs/fig_4_spatial.png}
\caption{\label{fig:spatial_omics}{}Morphology reveals the composition of tumour subregions. \textbf{Panel A.} Spatial map of cell types inferred from transcriptomics. \textbf{Panel B.} Morphological subregions obtained by Leiden clustering the image-derived feature embeddings. \textbf{Panel C.} Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type. Clusters 5-7 were merged due to similarity, cluster~10 removed due to low cell count (see Figure \ref{fig:spatial_c10}B).}
\end{figure}

These results demonstrate how LIB\_NAME can can be used outside traditional microscopy. When applied on extremely large images it produces biologically meaningful morphological clusters that validate and complement transcriptomic phenotypes. By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly add morphology to their spatial omics workflows, unlocking new insights into tissue organization.

\section{Discussion}
\label{sec:orgf37b369}
While point-and-click interfaces open the world of image analysis to many researchers, they are not as effective for computational workflows with no human-in-the-loop. In this work we introduced our new library LIB\_NAME, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks). It enables simpler automated profiling of microscopy data in short scripts and complex pipelines. The modularity it provides facilitates pipelines with better scaling capabilities for high-content microscopy, with or without cloud infrastructure.

The biologically interpretable features provided by LIB\_NAME complement deep learning features and offer a better mechanistic understanding of the underlying biology. A potential workflow would be to use deep learning features to cluster and then use LIB\_NAME features from these clusters. As a whole, when used in tandem with generalist image-processing tools, such as Cellpose for segmentation \cite{stringerCellposeGeneralistAlgorithm2021}, machine and deep learning workflows can be streamlined.

While developed for biology, engineered morphological features have proven valuable far beyond their original domain--from analysing microplastics in environmental samples \citep{ideharaExploringNileRed2025} to quality control in manufacturing \citep{ilhanEffectsProcessParameters2021} and agriculture automation \citep{xuDesignNonDestructiveSeed2024}. 
LIB\_NAME's lightweight, API-first design makes these powerful features accessible to any field working with image data, removing the barriers of biological-specific interfaces.

\section{Future work}
\label{sec:org5cdbb12}
We propose to make LIB\_NAME an imported dependency for CellProfiler, offering several benefits. It would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements. It would also mean that any changes made in LIB\_NAME propagate to CellProfiler, benefiting the community for which it was originally developed.

We also plan to develop a comprehensive test suite to guarantee mathematical correctness, which currently CellProfiler itself is lacking. A comprehensive test suite would enable more confident and expedient optimization for the most compute-consuming features, such as granularity, by providing rapid iteration capabilities. Once tests are in place, we could add support for just-in-time compiling and GPUs. We envision that LIB\_NAME could also be the place to develop and distribute new measurements from the community.

@@

\appendix
\setcounter{figure}{0}
% redefine \thefigure to prefix “A” and then use arabic
\renewcommand{\thefigure}{A\arabic{figure}}

% fix appendix section numbering
\setcounter{section}{5}
\renewcommand{\thesection}{\arabic{section}}
\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
\renewcommand{\thesubsubsection}{\thesubsection.\arabic{subsubsection}}

\label{sec:orgdd18dd8}
\subsection{Methods}
\label{sec:orgb3e9382}
\subsubsection{Data and software}
\label{sec:orgbda0ae2}
The codebase for LIB\_NAME is available on \url{https://anonymous.4open.science/anonymize/feature_extraction_image_profiling}. All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://anonymous.4open.science/r/2025_paper_feature_extraction_image_profiling/readme.org}. 
%The datasets we produced for this work are available on Zenodo. The selection of samples from JUMP and the images can be found on \url{https://zenodo.org/records/15390631/latest}. The image-based profiles produced with Cellprofiler and used for the comparisons in Figure 2 can be on \url{https://zenodo.org/records/15505477/latest}.
%

@@

The following figures show the processing steps for the spatial data used to generate Figure~\ref{fig:spatial_omics} in more detail. Raw data from two Xenium Spatial Transcriptomics available from the 10x Genomics Datasets homepage (name: "FFPE Human Breast with Custom Add-on Panel", release date: 2023-01-22) were downloaded and converted to the `SpatialData` format. The data comprises two samples, one infiltrating ductal carcinoma ("sample 1") and one invasive lobular carcinoma ("sample 2"). H\&E images from the same dataset were also downloaded, aligned to the respective reference images, and saved to the data object. These can be seen in Figure~\ref{fig:spatial_hne}A and Figure~\ref{fig:spatial_hne}B. A pattern of tessellating hexagons was generated for each sample, spanning the entire tissue section. These hexagons were subsetted to only those overlaying the specimens as visible in the H\&E image, resulting in 10321 hexagons for sample~1 and 8148 for sample~2 respectively. These hexagonal masks were used to automatically extract the underlying H\&E crops which were then featurized using `Squidpy`, resulting in 945 features each. These features were then pre-processed using `Pycytominer`, resulting in 252 features. We then removed batch effect using Harmony. UMAP projections of these features are shown in Figure~\ref{fig:spatial_umap}.

\begin{figure}[htbp]
\centering
\includegraphics[width=.75\linewidth]{./figs/fig_S1_spatial_umap.png}
\caption{\label{fig:spatial_umap}{}UMAP projection of the processed morphology features. \textbf{Panel A.} Colored by the sample of origin, we see good mixing after the Harmony integration. \textbf{Panel B.} Colored by Leiden clustering.}
\end{figure}

Due to the low cell count in Leiden cluster 10 for sample 2, as shown in Figure~\ref{fig:spatial_c10}B, the cluster was excluded for some analysis steps, and Figure~\ref{fig:spatial_omics}.

\begin{figure}[htbp]
\centering
\includegraphics[width=.7\linewidth]{./figs/fig_S4_spatial_c10.png}
\caption{\label{fig:spatial_c10}{}Barplots showing the total number of cells for each Leiden cluster in sample~1 (\textbf{A}) and sample~2 (\textbf{B}).}
\end{figure}

\newpage

Then, the gene expression data from the Xenium experiment was used to annotate cell types for both samples. For every Leiden cluster, as defined by the Leiden clustering of the joint morphology features from both samples, the cells covered by the respective hexagons were extracted. Figure~\ref{fig:spatial_hne} shows the raw H\&E images, the distribution of cell types across these, and the hexagons colored by their respective Leiden cluster. Of note is that the H\&E images both seemed to be partially corrupted, resulting in blacked-out sections. While the data contains cell segmentation masks for these sections, we filtered out hexagons overlaying these corrupted regions for the morphology featurization.

\begin{figure}[htbp]
\centering
\includegraphics[width=.65\linewidth]{./figs/fig_S2_spatial_hne.png}
\caption{\label{fig:spatial_hne}{}Overview about the spatial distribution of cell types and morphology clusters. \textbf{Panel A and B.} H\&E images of both samples. \textbf{Panel C and D.} Cell segmentation masks colored by cell type. \textbf{Panel E and F.} Hexagons colored by their morphology Leiden cluster.}
\end{figure}
